Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-03-04 msc-publication.tex
improve introduction, say more about the horizontal pressure gradient errors and how TF and cut cells proposed to reduce this error

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.

With terrain following (TF) layers the terrain's influence decays with height so that the bottommost layers follow the underlying surface closely while the uppermost layers are flat.  TF layers are usually implemented on a rectangular computational grid using transformed coordinates.

It is well-known that TF coordinates perform badly in the presence of steep orography: as model resolution increases, steeper gradients lead to larger truncation errors in calculating the horizontal pressure gradient which result in spurious winds \citep{dempsey-davis1998,steppeler2002}.

Much work has been done to reduce error associated with TF coordinates: firstly by smoothing the effects of terrain with height \citep{schaer2002,leuenberger2010,klemp2011} and, secondly, by improving the accuracy in calculating the horizontal pressure gradient itself \citep{klemp2011,zaengl2012}.

Despite their associated numerical errors, TF coordinates are attractive because their rectangular structure is simple to process by computer, boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}, and cell sizes remain almost constant \citep{jebens2011}.

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  This modification means that some cells become very small, which can reduce computational efficiency \citep{klein2009}, and several approaches have been tried to alleviate the problem \citep{steppeler2002,yamazaki-satomura2010,jebens2011}.

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  A comparison of TF and cut cells using real initial data by \citet{steppeler2006} found that precipitation patterns, temperature and wind fields were forecast more accurately in the cut cell model.

Many comparisons of TF and cut cell grids have made between different models.  On the contrary, this study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between BTF, SLEVE and cut cell grids for idealised, two-dimensional test cases from the literature.  \TODO{final bit of intro: in section 2, section 3 etc etc}

@@

Tracer magnitude and shape are well-preserved on all grids, both above the mountain at \(t = \SI{5000}{\second}\) and past the mountain at \(t = \SI{10000}{\second}\).  Advection is most accurate on the cut cell grid (figure~\ref{fig:advection-tracer}c) and regular grid (not shown).  As found by \citet{good2014}, the result is the same on both grids.  This is to be expected since the wind is zero in the region of the ground and both grids are orthogonal elsewhere.  On the BTF grid, the tracer is less distorted by the cubic upwind-biased scheme (figure~\ref{fig:advection-tracer}b) compared to the fourth-order centred scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}a).

>>>>

\section{Introduction}
Representing orography accurately in numerical weather prediction systems is necessary to model downslope winds and local precipitation patterns.  There are two main approaches to represent orography on a grid: terrain following layers and cut cells.  Both methods align cells in vertical columns.  Existing studies of cut cells have compared results with tests on terrain following grids from previous studies that used different models.  On the contrary, this study uses a single model to enable a like-for-like comparison between terrain following and cut cell grids.

With increasing model resolution, terrain can become steeper, making accurate calculation of the horizontal pressure gradient more difficult \citep{steppeler2002}.  Numerical errors in this calculation result in spurious winds and can cause numerical instability \citep{fast2003,webster2003}.  Terrain following and cut cell methods seek to reduce the error that is associated with steep orography.

With terrain following (TF) layers the terrain's influence decays with height so that the bottommost layers follow the underlying surface closely while the uppermost layers are flat.  There are two main approaches to minimizing errors associated with TF layers.  First, by smoothing the effects of terrain with height, the influence of the terrain is reduced, hence errors in the calculated horizontal pressure gradient are also reduced aloft \citep{schaer2002,leuenberger2010,klemp2011}.  However, the error is not reduced at the ground where steep terrain remains unmodified.

Second, numerical errors can also be reduced by improving the accuracy in calculating the horizontal pressure gradient itself.  TF layers are usually implemented using a coordinate transformation onto a rectangular computational domain, which introduces metric terms into the equations of motion.  The techniques proposed by \citet{klemp2011} and \cite{zaengl2012} both involve the calculation of the horizontal pressure gradient in the physical domain.  They found that, by avoiding the calculation of metric terms associated with the computational domain, numerical errors were reduced.

Despite their associated numerical errors, TF layers are in widespread operational use \citep{steppeler2003}.  They are attractive because their rectangular structure is simple to process by computer, boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}, and cell sizes remain almost constant \citep{jebens2011}.

Cut cells is an alternative method in which cells that lie entirely below the terrain are removed, and those that intersect the surface are modified in shape so that they more closely fit the terrain.  The resulting grid is orthogonal everywhere except where cells have been cut.  Hence, errors are still introduced when calculating the horizontal pressure gradient between cut and uncut cells.

The cut cell method can create some very small cells which reduce computational efficiency \citep{klein2009}, and several approaches have been tried to alleviate the problem \citep{steppeler2002,yamazaki-satomura2010,jebens2011}.

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  \TODO{anything else?}

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between BTF, SLEVE and cut cell grids for idealised, two-dimensional test cases from the literature.  \TODO{final bit of intro: in section 2, section 3 etc etc}

@@

Tracer magnitude and shape are well-preserved on all grids, both above the mountain at \(t = \SI{5000}{\second}\) and past the mountain at \(t = \SI{10000}{\second}\).  Advection is most accurate on the cut cell grid (figure~\ref{fig:advection-tracer}c) and regular grid (not shown).  As found by \citet{good2014}, the result is the same on both grids.  This is to be expected since the wind is zero in the region of the ground and flow aloft is aligned with the grids.  On the BTF grid, the tracer is less distorted by the cubic upwind-biased scheme (figure~\ref{fig:advection-tracer}b) compared to the fourth-order centred scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}a).
